---
title: "Decibels: What's the Reference?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Decibels: What's the Reference?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
author: Matthew Duggan
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The bioSNR package is an open-source SONAR equation calculator. The calculator is capable of handling simple to intermediate level acoustic problems associated with bioacoustics and passive acoustic monitoring (PAM) systems. 

This document gives quick examples of bioSNR's capabilities with dB operations. 

```{r setup, eval=FALSE}
#Stable - Install package from CRAN 
install.packages("bioSNR")

#Unstable - Install package from Github repository 
devtools::install_github("MattyD797/bioSNR")

#Attach package namespace to active libraries in Rstudio
library(bioSNR)
```

```{r setupR, eval=TRUE, include=FALSE}
library(bioSNR)
```


### Decibel

The logarithmic ratio of sound's power, intensity, or pressure between the sound source and the reference value. Which metric depends on the dB's reference of course, but the relationship is the same and adapted from the unit Bell or $Bel=\log_{10}(\frac{Power_2}{Power_1})$. Below are the specific equations for the three common metrics of sound (power, intensity, or pressure, respectively):

$$ dB\:re\:W_{ref}=10\log_{10}(\frac{W}{W_{ref}})  $$
$$ dB\:re\:I_{ref}=10\log_{10}(\frac{I}{I_{ref}})  $$
$$ dB\:re\:P_{ref}=10\log_{10}(\frac{P^2}{P_{ref}^2})\;\; |\;\;dB\:re\:P_{ref}=20\log_{10}(\frac{P}{P_{ref}})  $$

* $W$ is power in $W$.
* $I$ is intensity in $\frac{W}{m^2}$. 
* $P$ is the pressure in $\mu Pa$

The common reference levels that are always IMPORTANT to show are in the table below:

```{r referenceTable, echo=FALSE}
ref <- data.frame(row.names = c("Air", "Water"))
ref[,1] <- c("$20\\: \\mu Pa$", "$1\\: \\mu Pa$")
ref[,2] <- c("$1*10^{-12}\\: W/m^2$", "$6.7*10^{-19}\\: W/m^2$")
ref[,3] <- c("$1*10^{-12}\\: W$", "$6.7*10^{-19}\\: W$")
colnames(ref) <- c("Pressure", "Intensity", "Power")
t <- knitr::kable(ref,align=c(rep('c',times=3)), format = "html") 
kableExtra::kable_classic(t, full_width = FALSE) 
```
Since we are dealing in ratios, these are important relationships to keep in mind (notice how we need to increase dB by 10 to double the percieved 'loudness'):
```{r referenceTable2, echo=FALSE}
ref <- data.frame(row.names = c("1", "2", "3", "4", "5"))
ref[,1] <- c("dBs", "$+3$", "$+6$", "$+10$", "$+20$" )
ref[,2] <- c("Pressure", "$1.41*$", "$2*$", "$3.16*$", "$10*$" )
ref[,3] <- c("Intensity", "$2*$", "$4*$", "$10*$", "$100*$" )
ref[,4] <- c("Power", "$2*$", "$4*$", "$10*$", "$100*$" )
ref[,5] <- c("'Loudness'", "$1.23*$", "$1.52*$", "$2*$", "$4*$" )
t <- knitr::kable(ref,align=c(rep('c',times=5)), format = "html", col.names = NULL) 
kableExtra::kable_classic(t, full_width = FALSE) 
```

Simple operations with dBs have to be done in three steps of calculating linear values, add/subtract, and convert back to dB level. We can write out this operation as below:

1. $raw\;values = 10^{measurement/X}*ref\;level$
2. $Add/Subtract$
3. $dB\;value = X*log_{10}(measurement/ref\;level)$

* $X$ is 20 (pressure) or 10 (intensity, power)

Since comparing dBs can be more complicated than simple operations, the bioSNR makes adding, subtracting, and averaging relatively simple with a single function. 

#### Examples

1. Given some source level measurements you want to average them to get an idea on how many recorders you will need to record. Here are your measurements dB re $1\:\mu Pa$: 34, 57, 48, 66, 80, 55.

```{r opDB1}
opDB(c(34, 57, 48, 66, 80, 55), 20, op="avg")
```

* 67.087 dB re $1\:\mu Pa$

2. Given some source level measurements you want to add them because you have a lab assignment asking and just heard about bioSNR. Here are your measurements dB re $1*10^{-12} \frac{W}{m^2}$: 18, 27, 12, 22, 5, 50.

```{r opDB2}
opDB(c(18, 27, 12, 22, 5, 50), 10, op="add")
```

* 50.032 dB re $1*10^{-12} \frac{W}{m^2}$
