---
title: "bioSNR: An Introduction to the Physics Behind Bioacoustics"
output: rmarkdown::word_document
vignette: >
  %\VignetteIndexEntry{bioSNR: An Introduction to the Physics Behind Bioacoustics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
author: Matthew Duggan
nocite: '@*'
bibliography: ref_intro.bib 
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The bioSNR package is an open-source SONAR equation calculator. The calculator is capable of handling simple to intermediate level acoustic problems associated with bioacoustics and passive acoustic monitoring (PAM) systems. 

This document gives quick examples of bioSNR's more basic capabilities and functions while introducing basic physics behind sound. 

```{r setup, eval=FALSE}
#Stable - Install package from CRAN 
install.packages("bioSNR")

#Unstable - Install package from Github repository 
devtools::install_github("MattyD797/bioSNR")

#Attach package namespace to active libraries in Rstudio
library(bioSNR)
```

```{r setupR, eval=TRUE, include=FALSE}
library(bioSNR)
```

### Sound Wave: Frequency and Wavelength 

A sound wave is a longitudinal wave of particles that oscillate parallel to the wave front. This movement of particles results in a repeating pattern of *compression* and *refraction* from ambient pressure, resulting in a travelling pressure wave [@key3]. Along with the majority of organisms, our hearing is based on these pressure changes, but some species (mainly fish and squid) rely on the motion of the particles. Therefore, we typically only measure soundscapes through pressure changes.

Frequency (pitch) is the number of sound waves or oscillations produced in one second represented by the unit Hertz (Hz). In reference to our hearing sound with a frequency below 20 Hz is infrasound and sound above 20000 Hz is ultrasound. 

The relationship between the length of a wavelength and the frequency can be represented as: 

$$\lambda = \frac{c}{f}$$

* $\lambda$ is the wavelength (m)
* $c$ is speed of sound in $\frac{m}{s}$ 
* $f$ frequency in Hz (cycles per second). 

#### Examples

1. Marine animals like dolphins often use ultrasound signals for a range of important behaviors, such as communication. What is the wavelength of a false killer whale broadband click at 40 kHz?

```{r wof1}
#Find the wavelength (m) given a sound level of 75 kHz (75000 Hz) in WATER
wof(40000, c=15000)
```

* The wavelength is 0.375 meters. 

2. Bats produce ultrasonic sounds to find prey, such as the invasive mosquito. What is the frequency of a sheath-tailed bat echolocation that has a wavelength of 0.012?

```{r wof2}
#Find the frequency (Hz) given a wavelength of 0.015 in AIR
wof(0.012, c=350)
```

* The frequency is ~29 kHz. 

### Speed of Sound

The movement of a sound wave is similar in different medium, but mediums do have different physical properties that changes the speed of a wave. The speed of a wave is influenced by variable and dynamic changes of these properties, however, on average sound travels ~340 $\frac{m}{s}$ in air and ~1500 $\frac{m}{s}$ in salt water. In air, the factors which influence speed of sound are temperature, humidity, and wind speed. In water, the factors are temperature, depth, and salinity. 

The formula to find sound speed in _water_:

$$ c = 1449.2 + 4.6t – 0.055t^2 + 0.00029t^3 + (1.34-0.010t) * (s-35) + 0.0165z $$

* $c$ is sound speed in $\frac{m}{s}$
* $t$ is temperature in °C
* $s$ is salinity in parts per thousand (ppt)
* $z$ is depth in m

The formula to find sound speed in _air_:

$$ c = 331 + 0.6t $$

* $c$ is sound speed in $\frac{m}{s}$
* $t$ is temperature in °C

#### Example

1. How does the speed of airborne sound compare between Hawai'i (23°C) and a polar-like
region, such as Ithaca, NY (3°C)?

```{r soundSpeed1}
#Speed of sound in air at 30°C 
soundSpeed(23)
#Speed of sound in air at -10°C 
soundSpeed(3)
```
* In air, sound is 12 $\frac{m}{s}$ faster in 23°C compared to 3°C. 

2. How fast will a black durgon warning call travel in 30°C water with a salinity of 35 ppt and a depth of 10 meters?

```{r soundSpeed2}
#Speed of fish sound in water
soundSpeed(30, "water", 35, 10)
```
* In water, the call will travel at the speed of 1545.695 $\frac{m}{s}$. 

### Sound Reflection and Refraction

As a sound wave moves through a medium, it will be dynamically reflected and refracted. Reflection is like a ball bouncing off the ground in which the energy of a sound is directed into another direction by an object. Between the water-air boundary the critical angle is 15°. Refraction is when sound changes speed after crossing into another medium or boundary through a medium. Sound waves refract to the slower sound speed. 
The angle and longitudinal wave velocity of this new direction from reflection can be calculated utilizing Snell's law, where reflection angle is represented as $\Theta_1$ and refraction angle is represented as $\Theta_2$: 

$$ \frac{\sin\Theta_1}{{V_L}_1} = \frac{\sin\Theta_2}{{V_L}_2} $$

* ${V_L}_1$ is the longitudinal wave velocity in material 1 given in $\frac{m}{s}$
* ${V_L}_2$ is the longitudinal wave velocity in material 2 given in $\frac{m}{s}$

#### Examples

1. Calculate the reflection angle given the angle of a black-tailed godwit sound source is 64°, ${V_L}_1$ is 1564 $\frac{m}{s}$ and ${V_L}_2$ is 1494 $\frac{m}{s}$.

```{r Snells1}
snell(64,1564,1494)
```

* The reflection angle is 61.14°. 

2. Calculate the refraction angle given the angle of a Dixie Valley toad sound source is 15.5°, ${V_L}_1$ is 1493 $\frac{m}{s}$ and ${V_L}_2$ is 1502 $\frac{m}{s}$.

```{r Snells2}
snell(15.5,1493,1502)
```

* The refraction angle is 15.593°. 

### Sound Absorption, Impedance, Pressure, and Intensity


Sound absorption is the energy dissipation through a medium or the total transmission loss of sound from a source to a receiver. The transmission loss is impacted by the same factors impacting the speed of sound in each medium. Regarding absorption in air, the factors to take into consideration are the frequency of the sound source, ambient temperature, pressure, and relative humidity. In water, absorption is impacted by the frequency of the sound source, ambient temperature, depth, salinity and pH. Absorption is only one factor of transmission loss and more details on transmission loss are provided later, but for more information on the calculation please refer to **ISO 9613 Part 1** for sound absorption in air [@key2] and **Ainslie and McColm 1998** for sound absorption in water [@key1]. Note: There are several equations to estimate absorption in water for different ranges of predictors not functional in this package. 

Impedance is the measure of resistance of a medium to wave propagation, or in other words, the willingness of a sound wave to move through a medium [@key4]. The factors that affect impedance are the medium's properties and the type of wave propagation. The unit of specific acoustic impedance is the Rayl (Ry), which is equal to 1 $kg/(m^2s)$ or 1 $Ns/m^3$. For air, z= 428 Ry at
20°C and z = 413 Ry at 0°C. For water, z = 14.8 MRy. We can map the relationship of impedance to the speed of the sound and density of the medium: 

$$ z = c * p $$

* $z$ is the impedance in rayl or $\frac{kg}{m^2}$
* $c$ is the speed of sound in $\frac{m}{s}$
* $p$ is the density of the medium in $\frac{kg}{m^3}$

Sound pressure is simply the average deviation from the surrounding pressure of the medium. We can measure the sound pressure (pressure root mean square) from either simulated sine wave representations of sound waves $Prms = P_{zero2peak}/\sqrt{2}$ or typical by averaging the absolute pressure measurements $Prms = \sqrt{(P^2)_{average}}$. Below is a simulated sine wave representation to give you an understanding of $P_{zero2peak}$(blue) and $P_{peak2peak}$(red)[@key5]:

```{r sine graph, echo=FALSE, fig.align = 'center'}
#simulated sine sound wave
t=seq(0,4*pi,0.1)
y=2*sin(t)
s <- as.data.frame(cbind(t,y), ncol=2)

ggplot2::ggplot(s, ggplot2::aes(t,y))+
  ggplot2::geom_line()+
  ggplot2::geom_hline(yintercept=0, 
                      size=1) + 
  ggplot2::geom_segment(ggplot2::aes(x = pi/2, y = 0, xend = pi/2, yend = 2),
                  arrow = grid::arrow(length = grid::unit(0.25, "cm")), 
                  color = "blue") +
  ggplot2::geom_segment(ggplot2::aes(x = 5*pi/2, y = 0, xend = 5*pi/2, yend = 2),
                  arrow = grid::arrow(length = grid::unit(0.25, "cm")), 
                  color = "red") +
  ggplot2::geom_segment(ggplot2::aes(x = 5*pi/2, y = 0, xend = 5*pi/2, yend = -2),
                  arrow = grid::arrow(length = grid::unit(0.25, "cm")), 
                  color = "red") +
  ggplot2::theme(panel.grid.major = ggplot2::element_blank(), 
                 panel.grid.minor = ggplot2::element_blank(), 
                 panel.background = ggplot2::element_blank(), 
                 axis.text.x = ggplot2::element_blank(), 
                 axis.ticks.x = ggplot2::element_blank())+
  ggplot2::ylab(expression("Pressure (" * mu ~ "Pa)"))+
  ggplot2::xlab("")
  
```

Finally, the sound intensity can be measured by the relationship between pressure and the impedance. Therefore, the definition of intensity of sound is the average amount of energy transmitted per unit time through a unit area in a specified direction. The acoustic impedance of water is 3,500 times higher than in air, meaning sound in water is 3,500 times less intense compared to sound in air. On the contrary, a sound in water and air with the same intensity has a much smaller pressure in air [@key6;@key7]. The relaationship between impedence and pressure can be represented as:

$$ Sound \;intensity = sound \;pressure * particle \;velocity$$
$$ I = \frac{P_{rms}^2}{z}$$
* $I$ is the intensity of sound in $\frac{W}{m^2}$

#### Excersizes

1. A call of the rarest beaked whale, the spade-toothed beaked whale, has never been confirmed. What would be the absorption coefficient of a predicted call at 48kHz, a average salinity of 34ppt, a temperature of 18°C, a pH of 7.75, and estimated average depth of 2km?

```{r absorption1}
absorptionWater(48,7.75,18,34,2)
```

* The absorption of the call would be 9.434 $\frac{dB}{km}$. 

2. The land lobster, or _Dryococelus australis_, was a thought to be extinct insect in 1920, but was rediscovered in 2001. It is thought these insects make a small squeak noise when threatened. What would be the absorption coefficient with a call at 10 kHz, standard pressure, 20°C, and 99% humidity.   

```{r absorption2}
absorptionAir(10000,101.325,20,99)
```

* The absorption of the call would be 0.089 $\frac{dB}{m}$. 


3. A vector of hypothetical pressure values are given to you for a defensive call from a mangrove whiptail. The call was a 24 kHz made in the Indo-Pacific. What was the intensity of the sound?

```{r impedance}
#Assume standard impedance in water
z <- 14.8

#pressure values for example
press <- c(20, 24, 18, 34, 51, 29, 29, 15)

#pressure root mean square of pressure values
prms <- sqrt(mean(press^2))

#Formula for impedance
prms^2/z

```

* The intensity of the call would be 58.818 $\frac{W}{m^2}$. 

### References



